### 2020-03-26 HICSSSubmission/bp_discussion.tex
reducing text

Humans experience stress, focus and awakeness in different ways.  In
this paper, we have attempted to study these mental states in the
workplace using both participant self-reports and biometric data. We
discuss implications from our study for the workplace, including ways
in which the information might inform digitally-controlled or
digitally-informed parts of the workplace. We also discuss 
challenges imposed by the data and possible future paths of
research.

@@

Being able to accurately recognize periods of high-stress in knowledge
workers could enable more respectful workplaces. For example, an
ability to sense and predict stress in the moment could help companies
to prevent or de-escalate potentially dangerous situations in the
workplace, e.g. confrontation between co-workers. Building an 
understanding of stress and focus over time and in the moment
could also help create workplaces that are more conducive
to enabling knowledge workers to be more productive. For example,
this information could be used to feed an awareness dashboard of a team's stress level,
and avoid digital interruptions in high-stress or high-focus periods
similar to previous studies~\cite{zuger2017reducing}.  Building
an understanding of awakeness and focus could also enable the creation
of workplaces that are conducive to workers producing higher-quality
work. For example, if awakeness or focus decreases, they might 
be enhanced by adapting lighting in the workplace 
or scheduling breaks to prevent focus loss.

@@

As described in our study procedure,
due to our focus on stress, we offered participants an opportunity to be part of Thai Chi classes.
The two primary reasons for this intervention was to keep participants motivated to continue the self-reports over the long study period, and to offer a technique that might help reduce stress.
After consulting with other researchers, we decided to offer Tai Chi in the last month of the study, leaving the first month of the study unaltered. While participants were not required to take a Tai Chi class, they all took one per week except for two participants that opted out of the last two weeks of classes. Four participants explicitly stated that they liked the Tai Chi sessions or that they were ``excellent''.
These classes were performed once per week with a duration of one hour.

@@

Studying mental states, such as stress, awakeness and focus,
requires collecting a valid ground truth from each participant.
We spent considerable time when designing the study
 determining the exact questions to ask of participants,
consulting experts in the area, and basing
the questions and wording on previous research and studies. 
Despite the care taken, it is possible that the 
gathered data lacks reliability and validity. Some have
questioned the
reliability and validity of self-reports as we used in our study
due to subjective biases, lack of care in reporting, and the
highly individual nature of reporting aspects such as
stress~\cite{Hernandez11,Hovsepian15}.

@@

Study participants provided highly
imbalanced data in their survey responses, with most participants only
taking advantage of a subset of the Likert-scale values and the data
points mostly being clustered around the middle of the scale, as can
be seen in Table~\ref{responseDistribution}. While some of the
imbalance is expected due to certain classes, such as 'not stressed',
being more common in the workplace, this imbalance also provides
challenges in the training and assessment of a machine learning
classifier, as also found by others, e.g.~\cite{Exler16}. We addressed
this for the training by oversampling in case of few data samples for
the individual models and undersampling in case of a general model
where more data was available. Especially in light of this imbalance
in the data, the results we achieved with our models are
encouraging. For the assessment of the classifiers' performance we
addressed the imbalance by not just presenting accuracy, but also by
focusing on prediction and recall to examine the classifier's
performance in predicting the infrequent (yet more important) cases,
such as when a user is struggling to stay awake and an intervention or
warning might be needed most.
% \textbf{Unbalanced Data:}
% As shown in Table~\ref{responseDistribution}, the typical participant provided highly unbalanced survey responses, with most responses exhibiting a central tendency. Many participants provided zero or one very high (or very low) value on the Likert scale. In spite of this, our results show that by using biometric sensors and leveraging a machine learning approach, we are able to predict stress, focus, and awakeness better than the baseline, with improvements up to 84.64\%. Taking into account that the model correctly predicts many cases that happen infrequently, this is especially promising. These infrequent cases, such as when a user is highly stressed or struggling to stay awake, are in fact the most important cases to detect, as they are when intervention is needed most.

@@

Knowledge workers, a focus of our work and study, often spend
a large amount of time each day interacting with information on 
their computer at work. An interesting direction for future study
is to consider whether this computer interaction data, which can
be gathered non-invasively as work occurs, could serve to sense
and predict stress, focus and awakeness. Features that could be
investigated include keystrokes per minute, mouse clicks per minute
and changes in the active window title.

>>>>

\vspace{-4mm}
Humans experience stress, focus and awakeness in different ways.  In
this paper, we have attempted to study these mental states in the
workplace using both participant self-reports and biometric data. We
discuss implications from our study for the workplace, including ways
in which the information might inform digitally-controlled or
digitally-informed parts of the workplace. We also discuss 
challenges imposed by the data and possible future paths of
research.

@@

\vspace{-4mm}
Being able to accurately recognize periods of high-stress in knowledge
workers could enable more respectful workplaces. For example, an
ability to sense and predict stress in the moment could help companies
to prevent or de-escalate potentially dangerous situations in the
workplace, e.g. confrontation between co-workers. Building an 
understanding of stress and focus over time and in the moment
could also help create workplaces that are more conducive
to enabling knowledge workers to be more productive. For example,
this information could be used to feed an awareness dashboard of a team's stress level,
and avoid digital interruptions in high-stress or high-focus periods
similar to previous studies~\cite{zuger2017reducing}.  Building
an understanding of awakeness and focus could also enable the creation
of workplaces that are conducive to workers producing higher-quality
work. For example, if awakeness or focus decreases, they might 
be enhanced by adapting lighting in the workplace 
or scheduling breaks to prevent focus loss.

@@

\vspace{-4mm}
As described in our study procedure,
due to our focus on stress, we offered participants an opportunity to be part of Thai Chi classes.
The two primary reasons for this intervention was to keep participants motivated to continue the self-reports over the long study period, and to offer a technique that might help reduce stress.
After consulting with other researchers, we decided to offer Tai Chi in the last month of the study, leaving the first month of the study unaltered. While participants were not required to take a Tai Chi class, they all took one per week except for two participants that opted out of the last two weeks of classes. Four participants explicitly stated that they liked the Tai Chi sessions or that they were ``excellent''.
These classes were performed once per week with a duration of one hour.

@@

\vspace{-4mm}
Studying mental states, such as stress, awakeness and focus,
requires collecting a valid ground truth from each participant.
We spent considerable time when designing the study
 determining the exact questions to ask of participants,
consulting experts in the area, and basing
the questions and wording on previous research and studies. 
Despite the care taken, it is possible that the 
gathered data lacks reliability and validity. Some have
questioned the
reliability and validity of self-reports as we used in our study
due to subjective biases, lack of care in reporting, and the
highly individual nature of reporting aspects such as
stress~\cite{Hernandez11,Hovsepian15}.

@@

\vspace{-4mm}
Study participants provided highly
imbalanced data in their survey responses, with most participants only
taking advantage of a subset of the Likert-scale values and the data
points mostly being clustered around the middle of the scale, as can
be seen in Table~\ref{responseDistribution}. While some of the
imbalance is expected due to certain classes, such as 'not stressed',
being more common in the workplace, this imbalance also provides
challenges in the training and assessment of a machine learning
classifier, as also found by others, e.g.~\cite{Exler16}. We addressed
this for the training by oversampling in case of few data samples for
the individual models and undersampling in case of a general model
where more data was available. Especially in light of this imbalance
in the data, the results we achieved with our models are
encouraging. For the assessment of the classifiers' performance we
addressed the imbalance by not just presenting accuracy, but also by
focusing on prediction and recall to examine the classifier's
performance in predicting the infrequent (yet more important) cases,
such as when a user is struggling to stay awake and an intervention or
warning might be needed most.
% \textbf{Unbalanced Data:}
% As shown in Table~\ref{responseDistribution}, the typical participant provided highly unbalanced survey responses, with most responses exhibiting a central tendency. Many participants provided zero or one very high (or very low) value on the Likert scale. In spite of this, our results show that by using biometric sensors and leveraging a machine learning approach, we are able to predict stress, focus, and awakeness better than the baseline, with improvements up to 84.64\%. Taking into account that the model correctly predicts many cases that happen infrequently, this is especially promising. These infrequent cases, such as when a user is highly stressed or struggling to stay awake, are in fact the most important cases to detect, as they are when intervention is needed most.

@@

\vspace{-4mm}
Knowledge workers, a focus of our work and study, often spend
a large amount of time each day interacting with information on 
their computer at work. An interesting direction for future study
is to consider whether this computer interaction data, which can
be gathered non-invasively as work occurs, could serve to sense
and predict stress, focus and awakeness. Features that could be
investigated include keystrokes per minute, mouse clicks per minute
and changes in the active window title.

### 2020-03-26 HICSSSubmission/bp_intro_v4.tex
moved related work to the end, still need to cut 3 pages
reducing text

%!TEX root = bioPrediction_main.tex
\section{Introduction}
\textit{`The most valuable asset of a 21st-century institution (whether 
business or non-business) will be its knowledge workers and their 
productivity'}~\cite{drucker1999knowledge}. Knowledge workers constantly face 
challenges, such as a high work fragmentation, continuous disruptions and 
distractions, highly complex and demanding tasks, and long working 
hours~\cite{gonzalez2004constant,mark2008cost,czerwinski04diary}. 
These challenges, amongst others, can lead to stress in the workplace.
Stress is an ever-growing concern as it can lead to
fatigue, burnout and various other illnesses, ultimately resulting in work 
absences and marked productivity 
losses~\cite{hockey1997stress,setz2010stress,wrs2010}.

@@

We collected biometric data as well as
experience samples on stress, and two related human aspects: focus and
awakeness. We collected participant reports on focus because the
challenges that contribute to stress levels can also make it hard for
knowledge workers to stay focused and the level of focus
can affect
productivity~\cite{mark2014bored}.  We collected participant reports
on awakeness because a lack of awakeness (sleepiness) can result in
undesired consequences on work~\cite{connor02}.  The participants in
our study perform various job functions for a research and
development group within a single large corporation. For this study, they wore a novel, state-of-the-art
biometric armband sensor\footnote{Biovotion Everion
  sensor~\cite{everion}} with low invasiveness that captured heart-,
respiration- and skin-related measures. This modality was chosen to
ease longitudinal deployment in the field. Using machine learning, we
created classifiers and analyzed their ability to predict stress, focus, and awakeness levels. Considering all of these three human aspects helps us to consider whether one aspect might be easier to detect than another. If one aspect is easier to detect, it might serve
as a proxy or an indicator of the presence (or absence) of other aspects.

In our analysis, we identify several trends in day-to-day stress levels
that emerged over the course of our eight-week study.  The results of
our analysis also show that biometric signals collected from a single
minimally invasive sensor can be used to predict stress and its
related aspects accurately, with the abstract concept of focus
(predictably) being the hardest to detect. Our results further show
that knowledge workers' self-reported levels of stress, focus and
awakeness and their physiological manifestation and prediction can
vary substantially between individuals.
%Our results further show that overall combining biometric features with computer interaction information 
%results in an improvement
%over each of the training sets individually.
The main contributions of our work are:
\begin{itemize}
	\vspace{-3mm}
	\item A qualitative examination of how knowledge worker stress presents and fluctuates in the wild based on an eight-week field study with 14 office workers.
	\vspace{-3mm}
	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' stress, focus and awakeneess in the workplace based on a machine learning model trained on biometric data and experience samples.
	%\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.
	\vspace{-3mm}
	\item A discussion on the impact of applying this research to improve
the interaction of knowledge workers with their digital environment leading to an improvement in their productivity and well-being, besides a reflection on aspects that can be further improved in future studies.
    \end{itemize}

>>>>

%!TEX root = bioPrediction_main.tex
\section{Introduction}

\vspace{-4mm}
\textit{`The most valuable asset of a 21st-century institution (whether 
business or non-business) will be its knowledge workers and their 
productivity'}~\cite{drucker1999knowledge}. Knowledge workers constantly face 
challenges, such as a high work fragmentation, continuous disruptions and 
distractions, highly complex and demanding tasks, and long working 
hours~\cite{gonzalez2004constant,mark2008cost,czerwinski04diary}. 
These challenges, amongst others, can lead to stress in the workplace.
Stress is an ever-growing concern as it can lead to
fatigue, burnout and various other illnesses, ultimately resulting in work 
absences and marked productivity 
losses~\cite{hockey1997stress,setz2010stress,wrs2010}.

@@

We collected biometric data as well as
experience samples on stress, and two related human aspects: focus and
awakeness (i.e. wakefulness). We collected participant reports on focus because the
challenges that contribute to stress levels can also make it hard for
knowledge workers to stay focused and the level of focus
can affect
productivity~\cite{mark2014bored}.  We collected participant reports
on awakeness because a lack of awakeness (sleepiness) can result in
undesired consequences on work~\cite{connor02}.  The participants in
our study perform various job functions for a research and
development group within a single large corporation. For this study, they wore a novel, state-of-the-art
biometric armband sensor\footnote{Biovotion Everion
  sensor~\cite{everion}} with low invasiveness that captured heart-,
respiration- and skin-related measures. This modality was chosen to
ease longitudinal deployment in the field. Using machine learning, we
created classifiers and analyzed their ability to predict stress, focus, and awakeness levels. Considering all of these three human aspects helps us to consider whether one aspect might be easier to detect than another. If one aspect is easier to detect, it might serve
as a proxy or an indicator of the presence (or absence) of other aspects.

In our analysis, we identify several trends in day-to-day stress levels
that emerged over the course of our eight-week study.  The results of
our analysis also show that biometric signals collected from a single
minimally invasive sensor can be used to predict stress and its
related aspects accurately, with the abstract concept of focus
(predictably) being the hardest to detect. Our results further show
that knowledge workers' self-reported levels of stress, focus and
awakeness and their physiological manifestation and prediction can
vary substantially between individuals.
%Our results further show that overall combining biometric features with computer interaction information 
%results in an improvement
%over each of the training sets individually.
The main contributions of our work are:
\begin{itemize}[topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
	\item A qualitative examination of how knowledge worker stress presents and fluctuates in the wild based on an eight-week field study with 14 office workers.
	\item The creation and analysis of measures for the automatic monitoring of knowledge workers' stress, focus and awakeneess in the workplace based on a machine learning model trained on biometric data and experience samples.
	%\item A thorough analysis of ways to increase performance, including the training sample size, time window to collect data, and the inclusion of computer interaction features.
	\item A discussion on the impact of applying this research to improve
the interaction of knowledge workers with their digital environment leading to an improvement in their productivity and well-being, besides a reflection on aspects that can be further improved in future studies.
    \end{itemize}

### 2020-03-26 HICSSSubmission/bp_relatedWork.tex
reducing text

The study and prediction approach used in this paper is related
to previous studies of productivity-related indicators and studies
using biometrics to predict them.

@@

Our study focuses on the productivity-related indicators of stress, focus and awakeness.

@@

\vspace{-8mm}

@@

Much previous work relates to identifying and mitigating stress in an office environment.
Previous studies have measured stress by taking one of two possible approaches.
The first approach is to measure plasma catecholamine and cortisol as stress biomarkers~\cite{piazza10}.
This approach is impractical for use over prolonged periods of time, as in our eight-week study.
Further, this approach is imprecise because of the delay from the stress stimulation to the stress response, which 
may take from minutes to hours~\cite{Chandola10,Hellhammer09}.

@@

Focus refers to the allocation of limited cognitive processing resources~\cite{Anderson04}.
Mark et al.~\cite{mark2014bored} studied engagement in workplace activities by analyzing the digital activity of 32 information workers in situ for 5 days to understand how attentional states change with context.  They found that doing work that requires focus correlates with stress, while rote work correlates with happiness.

@@

Sleepiness (lack of awakeness) has been studied in the context of automobile 
accidents~\cite{connor02,Nordbakke07}.
These studies show a strong association between the level of acute driver sleepiness and the risk of injury crash.
Connor et al.~\cite{connor02} measure sleepiness using the Stanford sleepiness scale, similar to the seven-point Likert scale used in this study.

@@

Existing work~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015} analyzes a broad array of biometric signals and correlates them with individual's cognitive states and processes.
Zuger et al.~\cite{zuger18} used biometrics to sense interruptibility in the office~\cite{zuger18}.
Biometric signals have also been studied in the context of technology users. Parnin~\cite{Parnin2011} analyzes electromyography to measure sub-vocal utterances, and how these might be correlated with the programmer's perceived difficulty of programming tasks. Similarly, biometrics have been used to measure code difficulty~\cite{fritz2014using}
and using Near Infrared Spectroscopy to measure developer's cerebral blood flow~\cite{Nakagawa2014}.

>>>>

\vspace{-4mm}
The study and prediction approach used in this paper is related
to previous studies of productivity-related indicators and studies
using biometrics to predict them.

@@

\vspace{-4mm}
Our study focuses on the productivity-related indicators of stress, focus and awakeness.

@@

@@

\vspace{-8mm}
Much previous work relates to identifying and mitigating stress in an office environment.
Previous studies have measured stress by taking one of two possible approaches.
The first approach is to measure plasma catecholamine and cortisol as stress biomarkers~\cite{piazza10}.
This approach is impractical for use over prolonged periods of time, as in our eight-week study.
Further, this approach is imprecise because of the delay from the stress stimulation to the stress response, which 
may take from minutes to hours~\cite{Chandola10,Hellhammer09}.

@@

\vspace{-8mm}
Focus refers to the allocation of limited cognitive processing resources~\cite{Anderson04}.
Mark et al.~\cite{mark2014bored} studied engagement in workplace activities by analyzing the digital activity of 32 information workers in situ for 5 days to understand how attentional states change with context.  They found that doing work that requires focus correlates with stress, while rote work correlates with happiness.

@@

\vspace{-8mm}
Sleepiness (lack of awakeness) has been studied in the context of automobile 
accidents~\cite{connor02,Nordbakke07}.
These studies show a strong association between the level of acute driver sleepiness and the risk of injury crash.
Connor et al.~\cite{connor02} measure sleepiness using the Stanford sleepiness scale, similar to the seven-point Likert scale used in this study.

@@

\vspace{-4mm}
Existing work~\cite{sano2013stress,healey2005detecting,wijsman2011towards,zuger2015interruptibility,goyal2017intelligent,Parnin2011,Nakagawa2014,Radevski2015} analyzes a broad array of biometric signals and correlates them with individual's cognitive states and processes.
Zuger et al.~\cite{zuger18} used biometrics to sense interruptibility in the office~\cite{zuger18}.
Biometric signals have also been studied in the context of technology users. Parnin~\cite{Parnin2011} analyzes electromyography to measure sub-vocal utterances, and how these might be correlated with the programmer's perceived difficulty of programming tasks. Similarly, biometrics have been used to measure code difficulty~\cite{fritz2014using}
and using Near Infrared Spectroscopy to measure developer's cerebral blood flow~\cite{Nakagawa2014}.
